
# Using historical rainfall and incident data as a predictive damage estimation tool

Ben Smith 
January 2021 
REACH BGD Mission

The aim of this script is to generate an output displaying the recorded
daily rainfall for CXB against the incident data for the camps. This
should then be able to inform actors as to the likely result, and
therefore response, required for incoming rainfall. Once completed,
this should be presented to the IMAWG and the EPRWG for feedback and
dissemination.


### Preamble
Load in all of the data and packages:
```{r}
# install.packages("readxl")
library(readxl)
# install.packages("xlsx")

# install.packages("tidyr")
library(tidyr) # For rearranging the incident dataset 

# install.packages("dplyr")
library(dplyr) # For rearranging the incident dataset

# install.packages("plotly") # For interactive graphs
library(plotly)

# install.packages("knitr") # For outputting plotly in Rmd
# library(knitr) # this sets of a load of slow crashes...

# install.packages("zoo")
library(zoo)

df_incident <- read_xlsx("../../01 - Input Data/01_incident_reports/210916_SM_Daily_Incident_From2018_To2021-Sep14.xlsx")

# rename the column names to remove spaces:
colnames(df_incident) <- gsub(pattern = " ", replacement = "_", x = colnames(df_incident))


# Load in the rainfall data:
df_cxb_2019 <- read_xlsx("../../06 - Output and Published Datasets/reach_bgd_nathaz_tf_hydrometeorlogical_incidents_jan2019_-_jan_2020_v2.xlsx", sheet = "Database")

df_cxb_2020 <- read_xlsx("../../06 - Output and Published Datasets/REACH_BGD_NatHazTWG - Hydro-Meteorological Incident Database 2020.xlsx", sheet = "Hydro-Met Incident DB (daily)")

df_cxb_2021 = read.csv("../../02 - Reformatted Data/BMD Weather Data - 2021 - Cox Bazar.csv")

```

### Reformat the rainfall rainfall data:
Now reformat all of the data to smarten inconsistencies and to merge into a single rainfall dataset (df_rainfall).
```{r}  
# Convert 2019 to daily
df_cxb_2019 <- df_cxb_2019 %>% 
  mutate(Date = as.Date(Date_time, format = "%Y-%m-%d")) %>%
  select(Date, "Interval_GSB Coxs Bazaar-1227") %>%
  group_by(Date) %>%
  summarise(Daily_Rainfall = sum(`Interval_GSB Coxs Bazaar-1227`, na.rm = T)) %>%
  filter(format(Date, "%Y") != "2020") %>%
  mutate(Date = as.POSIXct(Date))

# Convert 2020 date to POSIXct, rainfall to numeric and remove mysterious duplicate:
df_cxb_2020$Date = as.POSIXct(df_cxb_2020$Date)
df_cxb_2020$Daily_Rainfall <- as.numeric(df_cxb_2020$`rain.GSB Cox's Bazaar-1227`)
df_cxb_2020 = df_cxb_2020[-which(duplicated(df_cxb_2020$Date)),]

# Convert 2021 to POSIXct and numeric:
df_cxb_2021$Date = as.POSIXct(as.Date(df_cxb_2021$`ï..Date`,  "%d/%m/%Y"))
df_cxb_2021$Daily_Rainfall = as.numeric(gsub(pattern = "Tr", 
                                             replacement = "0.5", 
                                             x = df_cxb_2021$`Rainfall..mm.`))

# Join all of the rainfall into a single dataframe
df_rainfall <- rbind(df_cxb_2019, 
                     df_cxb_2020 %>% select(Date, Daily_Rainfall),
                     df_cxb_2021 %>% select(Date, Daily_Rainfall))

```

### Check the Input Data
Vie the data if you wish to help understand the incident data then crop this down so that only the incidents relating to flooding are included in a df_flood dataframe.
```{r}
View(head(df_incident))

df_flood <- df_incident %>% 
  filter(Type_of_incident == "Flood") %>%
  select(Camp, Date_of_incident, Number_of_incidents, 
         Incident_Details, Affected_HHs, Affected_individuals, 'Casualties_(ind)',
         Injured_ind, Displaced_HH, Displaced_ind,
         Partially_Damaged_shelters, Totally_Damaged_shelters, Damaged_Shelter_2018,
         Damaged_waterpoints, Damaged_health_facilities,
         Damaged_food_distribution_site)
```

### Reformat the Incident Data to Daily
The incident data is currently noted on an incident by incident basis. For the first line of investigation we will group this by day so that we can see the combined affect of the rainfall accross all of the camps.
```{r}
df_flood_daily = df_flood %>% 
  select(-Incident_Details) %>%
  group_by(Date_of_incident) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))
  
# Aggregate the text columns into single entries per day, for later...
temp_text <- df_flood %>% 
  select(Date_of_incident, Camp, Incident_Details) %>%
  group_by(Date_of_incident) %>% 
  mutate(Camp = paste0(Camp, collapse = "; "),
         Incident_Details = paste0(Incident_Details, collapse = "; ")) %>%
  distinct()
  
df_flood_daily <- df_flood_daily %>% left_join(temp_text, by = "Date_of_incident") 

rm(temp_text)
```

### Join all of the data into a single dataframe
For ease, lets create a single master dataset with all of the rainfall and daily flood incident data.
```{r}
dates = data.frame(Date = seq.POSIXt(min(df_rainfall$Date[1],
                                         df_flood_daily$Date_of_incident[1]),
                                     max(tail(df_rainfall$Date, 1),
                                              tail(df_flood_daily$Date_of_incident, 1)),
                                     by = "day"))

df_master = dates %>% full_join(df_rainfall)
df_master = df_master %>% full_join(df_flood_daily, by = c("Date" = "Date_of_incident"))
```

### Create rainfall rolling averages
It is likely that many of the events will be due to multiple days of rainfall
Therefore, create some rolling averages to check this.
```{r}
moving_sum = function(x, n){y = rollsum(x, k = n, align = "right", fill = NA, na.rm = T)}

df_master <- df_master %>% 
  mutate(
    "day2" = moving_sum(df_master$Daily_Rainfall, n=2),
    "day3" = moving_sum(df_master$Daily_Rainfall, n=3),
    "day4" = moving_sum(df_master$Daily_Rainfall, n=4),
    "day5" = moving_sum(df_master$Daily_Rainfall, n=5)) %>%
  relocate(c("day2", "day3", "day4", "day5"), .after = "Daily_Rainfall")

# "day3" = rollsum(df_master$Daily_Rainfall, k = 3, align = "right", fill = NA) 
# ^^^ This discounts values with NAs in their window.
```

### Initial Graphing
Let's set up some plots to get an idea of how the incident and rainfall data compare.

First, let's ;look at the incidents (number of individuals affected) over time:
```{r}
# Plot the affected individuals from events:
plot_ly(x = df_master$Date, 
        y = df_master$Affected_individuals,
        type = 'scatter', mode = 'lines+markers',
        hovertext = df_master$Incident_Details)
```

Let's then plot when the incidents occurred alongside the rainfall. Incidents are shown in blue below the xaxis.
```{r}
vlines = data.frame(
  x = c(t(data.frame(a = df_master$Date[which(df_master$Affected_individuals>0)], 
                     b = df_master$Date[which(df_master$Affected_individuals>0)],
                     c = rep(NA, length(which(df_master$Affected_individuals>0)))))),
  y = c(matrix(rep(c(1,-20,NA), 
                   length=3*length(which(df_master$Affected_individuals>0))),
               nrow=3)))

plot_ly(x = vlines$x, y = vlines$y,
        type = "scatter", mode = "lines",
        name = "Flooding_incidents") %>%
  add_trace(x = df_master$Date, y = df_master$Daily_Rainfall,
            mode = "lines", type = "scatter",
            name = "Rainfall")
```

We can also see below how the different consequences compare:
```{r}
# Plot a range of consequences from daily rainfall: 
plot_ly(x = df_master$Daily_Rainfall, y = df_master$Number_of_incidents,
        type = "scatter", mode = "markers", name = "Number_of_incidents") %>%
  add_trace(x = df_master$Daily_Rainfall, y = df_master$Affected_HHs,
            type = "scatter", mode = "markers", name = "Affected_HHs") %>%
  add_trace(x = df_master$Daily_Rainfall, y = df_master$Affected_individuals,
            type = "scatter", mode = "markers", name = "Affected_individuals") %>%
  add_trace(x = df_master$Daily_Rainfall, y = df_master$Partially_Damaged_shelters,
            type = "scatter", mode = "markers", name = "Partially_Damaged_shelters") %>%
  layout(xaxis = list(title = "Daily Rainfall (mm)"),
         yaxis = list(title = "Recorded Consequence"))


# Check whether affected households and individuals correlate:
View(round(cor(x = df_master[, 2:19], y = NULL, # [df_master$Affected_HHs>0, 2:19]
               method = "pearson", use = "pairwise.complete.obs"), 2))

plot_ly(x = df_master$Affected_individuals, y = df_master$Daily_Rainfall,
        type = "scatter", mode = "markers", name = "1") %>%
  add_trace(x = df_master$Affected_individuals, y = df_master$day2,
          type = "scatter", mode = "markers", name = "2") %>%
  add_trace(x = df_master$Affected_individuals, y = df_master$day3,
            type = "scatter", mode = "markers", name = "3") %>%
  add_trace(x = df_master$Affected_individuals, y = df_master$day4,
            type = "scatter", mode = "markers", name = "4") %>%
  add_trace(x = df_master$Affected_individuals, y = df_master$day5,
            type = "scatter", mode = "markers", name = "5") %>%
  layout(yaxis = list(title = "Daily Rainfall (mm)"),
         xaxis = list(title = "Recorded Consequence"))
```

### Create Output


### Caveats and Notes

```{r}
# --- Number of recordings:
# There are a large number of incident records that do not have rainfall data:
length(
  which(
    is.na(
      df_master$Daily_Rainfall[which(df_master$Affected_individuals>0)]
      )
    )
  )
# >>> 45 missing values out of 86, largely in 2018.
    # Though 2020 has a lot of missing data too.
```

### Unused Code
```{r}
  add_trace(x = df_master$Date_of_incident,
            y = df_master$Affected_individuals,
            type = 'scatter', mode = 'markers',
            name = "Incidents", yaxis = "y2",
            hovertext = df_master$Incident_Details) %>%
  layout(yaxis2 = list(side = "right",  overlaying = "y", sharey = T))
```


